---
title: "Assignment 6: Build and Evaluate Tree Models"
subtitle: "SoarDDDS8555-6 - Main Controller"
author: "Soar, Donald"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(reticulate)

python <- "C:/Users/Soar/Documents/.virtualenvs/SoarDDDS8555-6/Scripts/python.exe"

use_python(
  python,
  required = TRUE
)

```


```{r}


dir.create("data", showWarnings = FALSE)

dir.create(
  "scripts",
  showWarnings = FALSE
)

dir.create(
  "output/tables",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "output/figures",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "output/results",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "output/submissions",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "output/bart",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "output/logistic",
  recursive = TRUE,
  showWarnings = FALSE
)

```


```{r}

system2(
  command = "python",
  args = "scripts/assignment-6-question-1.py"
)

```



```{r}

reticulate::source_python(
  "scripts/01_inspect_data.py"
)

reticulate::source_python(
  "scripts/02_eda.py"
)

reticulate::source_python(
  "scripts/03_bagging.py"
)

reticulate::source_python(
  "scripts/04_decision_tree.py"
)

reticulate::source_python(
  "scripts/05_random_forest.py"
)

reticulate::source_python(
  "scripts/06_boosting.py"
)

reticulate::source_python(
  "scripts/07_logistic_regression.py"
)

# BART intentionally remains separate because of its
# substantially greater computational cost.

system2(
  command = "python",
  args = "scripts/08_bart.py"
 )

```



```{r}

 reticulate::source_python(
  "scripts/09_decision_tree_kaggle_submission.py"
 )


```


```{r}
 reticulate::source_python(
   "scripts/10_bagging_kaggle_submission.py"
 )

```


```{r}
 reticulate::source_python(
   "scripts/11_random_forest_kaggle_submission.py"
 )

```


```{r}

 reticulate::source_python(
   "scripts/12_boosting_kaggle_submission.py"
 )

```


```{r}

reticulate::source_python(
 "scripts/L2_Regularized_Multinomial_Logistic_Regression.py"
)

```



```{r}

 reticulate::source_python(
   "scripts/13_model_comparison.py"
 )

```


```{r}
# system2(
#   command = "python",
#   args = "-m pip install numpy pandas scikit-learn pymc pymc-bart"
# )

 system2(
    command = "python",
    args = "scripts/14_bart_validation_andevaluation.py"
 )

```


```{r}
 system2(
   command = "python",
   args = "scripts/15_bart_kaggle_submission.py"
 )

```



```{r}

rmarkdown::render(
   "SoarDDDS8555-6-Report.Rmd",
   output_format = "html_document"
)

```


```{r}

 rmarkdown::render(
   "SoarDDDS8555-6-Report.Rmd",
   output_format = "pdf_document"
 )

```



# References

James, G., Witten, D., Hastie, T., Tibshirani, R., & Taylor, J. (2023). An introduction to statistical learning: With applications in Python. Springer.

Breiman, L. (2001). Random forests. Machine Learning, 45(1), 5–32. https://doi.org/10.1023/A:1010933404324

Ganie, S. M., Dutta Pramanik, P. K., & Zhao, Z. (2025). Lifestyle data-based multiclass obesity prediction with interpretable ensemble models incorporating SHAP and LIME analysis. Scientific Reports, 15, 36916. https://doi.org/10.1038/s41598-025-20936-4

Chipman, H. A., George, E. I., & McCulloch, R. E. (2010). BART: Bayesian additive regression trees. The Annals of Applied Statistics, 4(1), 266–298. https://doi.org/10.1214/09-AOAS285

